Probability Aggregation: Dynamic Hierarchical Modeling of Sparse Expert Beliefs

### 2013-09-01 paper.tex
Edits (some AOAS guidelines)

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting. In response to the lack of appropriate modeling procedures, this paper introduces a time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is highly interpretable and can be used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

The second matter is related to the distribution of the class labels in the data. If the set of occurred events is much larger than the set of events that did not occur (or \textit{vice versa}), the dataset is called \textit{imbalanced}. On such data the modeling procedure can end up over-focusing on the larger class, and as a result, give very accurate forecast performance over the larger class at the cost of performing poorly over the smaller class (see, e.g., \citet{chen2009learning, wallace2012class}). Fortunately, it is often possible to use a well-balanced version of the data. The first step is to find a partition $S_0$ and $S_1$ of the question indices $\{1, 2, \dots, K\}$ such that the equality $\sum_{k \in S_0} T_k  = \sum_{k \in S_1} T_k$ is as closely approximated as possible. This is equivalent to an NP-hard problem known in computer science as the \textit{Partition Problem}:  determine whether a given set of positive integers can be partitioned into two sets such that the sums of the two sets equal to each other (see, e.g., \citet{karmarkar1982differencing, hayes2002easiest}). A simple solution is to use a greedy algorithm that iterates through the values of $T_k$ in descending order, assigning each $T_k$ to the subset that currently has the smaller sum (see, e.g. \citet{kellerer2004knapsack, gent1996phase} for more details on the Partition Problem). After finding a well-balanced partition, the next step is to assign the class labels such that the labels for the questions in $S_x$ are equal to $x$ for $x = 0$ or $1$. Recall from section \ref{calibration_step} that $Z_k$ represents the event indicator associated with the $k$th question. To define a balanced set of indicators $ \tilde{Z}_k$ for all $k \in S_x$, let
\begin{align*}
 \tilde{Z}_k &= x\\
\tilde{p}_{i,t,k} &=  \begin{cases} 
1-p_{i,t,k} & \text{if } Z_k = 1-x\\
p_{i,t,k}& \text{if } Z_k = x,
\end{cases}
\end{align*}
where $i = 1, \dots, I_k$, and $t = 1, \dots, T_k$. The resulting set $$\left\{\left(\tilde{Z}_k, \{\tilde{p}_{i,t,k} | i = 1, \dots, I_k, t = 1, \dots, T_k \}\right) \right\}_{k=1}^K$$ is a balanced version of the data. 
This procedure was used to balance our real-world dataset both in terms of events and time points. The final output splits the events exactly in half ($|S_0| = |S_1| = 83$) such that number of time points in the first and second halves are 8,737 and 8,738, respectively.

\subsection{Out-of-Sample Forecasting}
\label{forecasting}
This section evaluates the out-of-sample predictive performance of \textit{Sample-Then-Calibrate} against several other probability aggregation procedures. The models are allowed to utilize a training set before making predictions on an independent testing set. In order to clarify some of the upcoming notation, let $S_{train}$ and $S_{test}$ be index sets that partition the data into training and testing sets of sizes $|S_{train}| = N_{train}$ and $|S_{test}| = 166 - N_{test}$, respectively. This means that the $k$th question is in the training set if and only if $k \in S_{train}$. The competing models are as follows.

@@

\item \textit{Exponentially Weighted Moving Average (EWMA)}. If 
\begin{eqnarray*}
\bar{p}_{t,k}  = \frac{1}{N_{t,k}} \sum_{i=1}^{N_{t,k}} p_{i,t,k},
\end{eqnarray*}
is the average probability forecast given at time $t$ for the $k$th question, then the EWMA forecasts for the $k$th problem are obtained recursively from
\begin{eqnarray*}
\hat{p}_{t,k}(\alpha) =
\begin{cases}
 \bar{p}_{1,k} & \text{ for } t  = 1 \\
\alpha \bar{p}_{t,k}  + (1-\alpha) \hat{p}_{t-1,k}(\alpha) & \text{ for } t > 1
\end{cases}
\end{eqnarray*}
where the input parameter $\alpha$ is learned from the training set by 
\begin{eqnarray*}
\hat{\alpha} &=& \argmin_{\alpha \in [0,1]} \sum_{k \in S_{train}} \sum_{t=1}^{T_k} \left(Z_k - \hat{p}_{t,k}(\alpha) \right)^2
\end{eqnarray*}

\item  \textit{Exponentially Weighted Moving Logit Aggregator (EWMLA)}. This is a moving version of the aggregator $\hat{p}_G(\boldsymbol{b})$ that was introduced in \cite{satopaa}. If $\boldsymbol{p}_{t,k}$ is a vector collecting all the probability forecasts made for the $k$th question at time $t$, then the EWMLA forecasts are found recursively from
\begin{eqnarray*}
\hat{p}_{t,k}(\alpha, \boldsymbol{b}) =
\begin{cases}
G_{t,k}(\boldsymbol{b} ) & \text{ for } t  = 1 \\
\alpha G_{t,k}(\boldsymbol{b} )  + (1-\alpha) \hat{p}_{t-1,k}(\alpha, \boldsymbol{b})  & \text{ for } t > 1
\end{cases}
\end{eqnarray*}
where
\begin{eqnarray*}
G_{t,k}(\nu )  &=& \left( \prod\limits_{i=1}^{N_{t,k}} \left( \frac{p_{i, t, k}}{1-p_{i, t, k}} \right)^{ \frac{\boldsymbol{e}_{i,k}'\boldsymbol{b}}{N_{t,k}}} \right) \Bigg/ \left(1+ \prod\limits_{i=1}^{N_{t,k}} \left( \frac{p_{i, t, k}}{1-p_{i, t, k}} \right)^{\frac{\boldsymbol{e}_{i,k}' \boldsymbol{b}}{N_{t,k}}} \right)
\end{eqnarray*}
The vector $\boldsymbol{b}$ collects the bias terms of the different expertise groups. Therefore it is equivalent to the bias vector found under \textit{Sample-Then-Calibrate}. The term $\boldsymbol{e}_{i,k}$ is a vector of length 5 indicating which level of self-reported expertise the $i$th expert in the $k$th question belongs to. For instance, if $\boldsymbol{e}_{i,k} = [0, 1, 0, 0, 0]$, then the expert identifies himself with the expertise level two. The tuning parameters $(\alpha, \boldsymbol{b})$ are learned from the training set by
\begin{eqnarray*}
(\hat{\alpha}, \hat{ \boldsymbol{b}}) &=& \argmin_{ \boldsymbol{b} \in \R^5, \alpha \in [0,1]} \sum_{k \in S_{train}} \sum_{t=1}^{T_k} \left(Z_k - \hat{p}_{t,k}(\alpha,  \boldsymbol{b})\right)^2
\end{eqnarray*}

@@

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on expert probability forecasts. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample settings. This has direct implications  on predictive power and model interpretability. Firstly, the model was shown to outperform other probability aggregators in terms of forecasting ability. In particular, the model was deemed a well-balanced compromise that avoids over-fitting without being overly conservative. Secondly, the crowd belief was used as the no-bias reference point to study the bias among groups of experts with different levels of self-reported expertise. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific parameters. These quantities have clear interpretations and can be combined in many interesting ways to explore a range of hypotheses about different types of questions and expert behavior.

>>>>

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting. In response to the lack of appropriate modeling procedures, this paper introduces a time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is highly interpretable and can be used for:
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

The second matter is related to the distribution of the class labels in the data. If the set of occurred events is much larger than the set of events that did not occur (or \textit{vice versa}), the dataset is called \textit{imbalanced}. On such data the modeling procedure can end up over-focusing on the larger class, and as a result, give very accurate forecast performance over the larger class at the cost of performing poorly over the smaller class (see, e.g., \citet{chen2009learning, wallace2012class}). Fortunately, it is often possible to use a well-balanced version of the data. The first step is to find a partition $S_0$ and $S_1$ of the question indices $\{1, 2, \dots, K\}$ such that the equality $\sum_{k \in S_0} T_k  = \sum_{k \in S_1} T_k$ is as closely approximated as possible. This is equivalent to an NP-hard problem known in computer science as the \textit{Partition Problem}:  determine whether a given set of positive integers can be partitioned into two sets such that the sums of the two sets equal to each other (see, e.g., \citet{karmarkar1982differencing, hayes2002easiest}). A simple solution is to use a greedy algorithm that iterates through the values of $T_k$ in descending order, assigning each $T_k$ to the subset that currently has the smaller sum (see, e.g. \citet{kellerer2004knapsack, gent1996phase} for more details on the Partition Problem). After finding a well-balanced partition, the next step is to assign the class labels such that the labels for the questions in $S_x$ are equal to $x$ for $x = 0$ or $1$. Recall from section \ref{calibration_step} that $Z_k$ represents the event indicator associated with the $k$th question. To define a balanced set of indicators $ \tilde{Z}_k$ for all $k \in S_x$, let
\begin{align*}
 \tilde{Z}_k &= x\\
\tilde{p}_{i,t,k} &=  \begin{cases} 
1-p_{i,t,k}, & \text{if } Z_k = 1-x,\\
p_{i,t,k}, & \text{if } Z_k = x,
\end{cases}
\end{align*}
where $i = 1, \dots, I_k$, and $t = 1, \dots, T_k$. The resulting set $$\left\{\left(\tilde{Z}_k, \{\tilde{p}_{i,t,k} | i = 1, \dots, I_k, t = 1, \dots, T_k \}\right) \right\}_{k=1}^K$$ is a balanced version of the data. 
This procedure was used to balance our real-world dataset both in terms of events and time points. The final output splits the events exactly in half ($|S_0| = |S_1| = 83$) such that number of time points in the first and second halves are 8,737 and 8,738, respectively.

\subsection{Out-of-Sample Forecasting}
\label{forecasting}
This section evaluates the out-of-sample predictive performance of \textit{Sample-Then-Calibrate} against several other probability aggregation procedures. The models are allowed to utilize a training set before making predictions on an independent testing set. In order to clarify some of the upcoming notation, let $S_{train}$ and $S_{test}$ be index sets that partition the data into training and testing sets of sizes $|S_{train}| = N_{train}$ and $|S_{test}| = 166 - N_{test}$, respectively. This means that the $k$th question is in the training set if and only if $k \in S_{train}$. The competing models are as follows:

@@

\item \textit{Exponentially Weighted Moving Average (EWMA)}. If 
\begin{eqnarray*}
\bar{p}_{t,k}  = \frac{1}{N_{t,k}} \sum_{i=1}^{N_{t,k}} p_{i,t,k},
\end{eqnarray*}
is the average probability forecast given at time $t$ for the $k$th question, then the EWMA forecasts for the $k$th problem are obtained recursively from
\begin{eqnarray*}
\hat{p}_{t,k}(\alpha) =
\begin{cases}
 \bar{p}_{1,k}, & \text{ for } t  = 1, \\
\alpha \bar{p}_{t,k}  + (1-\alpha) \hat{p}_{t-1,k}(\alpha), & \text{ for } t > 1,
\end{cases}
\end{eqnarray*}
where the input parameter $\alpha$ is learned from the training set by 
\begin{eqnarray*}
\hat{\alpha} &=& \argmin_{\alpha \in [0,1]} \sum_{k \in S_{train}} \sum_{t=1}^{T_k} \left(Z_k - \hat{p}_{t,k}(\alpha) \right)^2
\end{eqnarray*}

\item  \textit{Exponentially Weighted Moving Logit Aggregator (EWMLA)}. This is a moving version of the aggregator $\hat{p}_G(\boldsymbol{b})$ that was introduced in \cite{satopaa}. If $\boldsymbol{p}_{t,k}$ is a vector collecting all the probability forecasts made for the $k$th question at time $t$, then the EWMLA forecasts are found recursively from
\begin{eqnarray*}
\hat{p}_{t,k}(\alpha, \boldsymbol{b}) =
\begin{cases}
G_{t,k}(\boldsymbol{b} ), & \text{ for } t  = 1, \\
\alpha G_{t,k}(\boldsymbol{b} )  + (1-\alpha) \hat{p}_{t-1,k}(\alpha, \boldsymbol{b}),  & \text{ for } t > 1,
\end{cases}
\end{eqnarray*}
where
\begin{eqnarray*}
G_{t,k}(\nu )  &=& \left( \prod\limits_{i=1}^{N_{t,k}} \left( \frac{p_{i, t, k}}{1-p_{i, t, k}} \right)^{ \frac{\boldsymbol{e}_{i,k}'\boldsymbol{b}}{N_{t,k}}} \right) \Bigg/ \left(1+ \prod\limits_{i=1}^{N_{t,k}} \left( \frac{p_{i, t, k}}{1-p_{i, t, k}} \right)^{\frac{\boldsymbol{e}_{i,k}' \boldsymbol{b}}{N_{t,k}}} \right)
\end{eqnarray*}
The vector $\boldsymbol{b}$ collects the bias terms of the different expertise groups. Therefore it is equivalent to the bias vector found under \textit{Sample-Then-Calibrate}. The term $\boldsymbol{e}_{i,k}$ is a vector of length 5 indicating which level of self-reported expertise the $i$th expert in the $k$th question belongs to. For instance, if $\boldsymbol{e}_{i,k} = [0, 1, 0, 0, 0]$, then the expert identifies himself with the expertise level two. The tuning parameters $(\alpha, \boldsymbol{b})$ are learned from the training set by
\begin{eqnarray*}
(\hat{\alpha}, \hat{ \boldsymbol{b}}) &=& \argmin_{ \boldsymbol{b} \in \R^5, \alpha \in [0,1]} \sum_{k \in S_{train}} \sum_{t=1}^{T_k} \left(Z_k - \hat{p}_{t,k}(\alpha,  \boldsymbol{b})\right)^2
\end{eqnarray*}

@@

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on expert probability forecasts. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample settings. This has direct implications  on predictive power and model interpretability. First, the model was shown to outperform other probability aggregators in terms of forecasting ability. In particular, the model was deemed a well-balanced compromise that avoids over-fitting without being overly conservative. Second, the crowd belief was used as the no-bias reference point to study the bias among groups of experts with different levels of self-reported expertise. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific parameters. These quantities have clear interpretations and can be combined in many interesting ways to explore a range of hypotheses about different types of questions and expert behavior.
